---
title: "TENxIO: Import Single Cell Data Files"
author: "Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TENxIO Quick Start Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
Package: TENxIO
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Installation

```{r,eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("waldronlab/TENxIO")
```

# Load the package

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(TENxIO)
```

# Description

`TENxIO` offers an set of classes that allow users to easily work with files
typically obtained from the 10X Genomics website. These are commonly outputs
of the Cell Ranger pipeline.

# Supported Formats

| **Extension**       | **Class**     | **Imported as**      |
|---------------------|---------------|----------------------|
| .h5                 | TENxH5        | SingleCellExperiment |
| .mtx / .mtx.gz      | TENxMTX       | dgCMatrix            |
| .tar.gz             | TENxFileList  | SingleCellExperiment |
| peak_annotation.tsv | TENxPeaks     | GRanges              |
| fragments.tsv.gz    | TENxFragments | RaggedExperiment     |
| .tsv / .tsv.gz      | TSVFile\*     | tibble               |

**Note** (\*). The `TSVFile` class is used internally and not exported.

# TENxFile

The `TENxFile` class is the catch-all class superclass that allows transition
to subclasses pertinent to specific files. It inherits from the `BiocFile`
class and allows for easy dispatching `import` methods.

```{r}
showClass("TENxFile")
```

## `ExperimentHub` resources

`TENxFile` can handle resources from `ExperimentHub` with careful inputs. 
For example, one can import a `TENxBrainData` dataset via the appropriate
`ExperimentHub` identifier (`EH1039`):

```{r}
hub <- ExperimentHub::ExperimentHub()
hub["EH1039"]
```

Currently, `ExperimentHub` resources do not have an extension and it is best to
provide that to the `TENxFile` constructor function.

```{r}
fname <- hub[["EH1039"]]
TENxFile(fname, extension = "h5", group = "mm10", version = "2")
```

## TENxH5

One of the main data formats provided by the 10X website are HDF5 files.
To import those files, we use the TENxH5 contructor function.

```{r}
h5f <- system.file(
    "extdata", "pbmc_granulocyte_ff_bc_ex.h5",
    package = "TENxIO", mustWork = TRUE
) 
h5f
```

Here we provide a bespoke show method for such files given that we have some
idea of the structure to expect within these files:

```{r}
con <- TENxH5(h5f)
con
```

### import TENxH5 method

We can simply use the import method to convert the file representation to
a Bioconductor class representation, typically a `SingleCellExperiment`.

```{r}
import(con)
```

**Note**. Future versions of the package could support alternative
representations to `SingleCellExperiment`

# TENxMTX

Matrix Market formats are also supported (`.mtx` extension). These are typically
imported as SummarizedExperiment as they usually contain count data.

```{r}
mtxf <- system.file(
    "extdata", "pbmc_3k_ff_bc_ex.mtx",
    package = "TENxIO", mustWork = TRUE
)
con <- TENxMTX(mtxf)
con
import(con)
```

# TENxFileList

The `TENxFileList` class easily allows importing multiple files within a
`tar.gz` archive. The `untar` function can list the files compressed within the
tarball.

```{r}
fl <- system.file(
    "extdata", "pbmc_granulocyte_sorted_3k_ff_bc_ex_matrix.tar.gz",
    package = "TENxIO", mustWork = TRUE
)
untar(fl, list = TRUE)
```

Using a similar import process accross all file types, one can easily obtain
a Bioconductor representation that is ready for analysis. `TENxFileList` can be
imported to `SingleCellExperiment`.

```{r}
con <- TENxFileList(fl)
import(con)
```

# TENxPeaks

Peak files can be handled with the `TENxPeaks` class. These files are usually
named `*peak_annotation` files with a `.tsv` extension. Peak files are
represented as `GRanges`.

```{r}
pfl <- system.file(
    "extdata", "pbmc_granulocyte_sorted_3k_ex_atac_peak_annotation.tsv",
    package = "TENxIO", mustWork = TRUE
)
tenxp <- TENxPeaks(pfl)
peak_anno <- import(tenxp)
peak_anno
```

# TENxFragments

Fragment files are quite large and we make use of the `Rsamtools` package to
import them with the `yieldSize` parameter. By default, we use a `yieldSize` of
200.

```{r}
fr <- system.file(
    "extdata", "pbmc_3k_atac_ex_fragments.tsv.gz",
    package = "TENxIO", mustWork = TRUE
)
```

Internally, we use the `TabixFile` constructor function to work with indexed
`tsv.gz` files.

**Note**. A warning is emitted whenever a `yieldSize` parameter is not set.

```{r}
tfr <- TENxFragments(fr)
tfr
```

Because there may be a variable number of fragments per barcode, we use a
`RaggedExperiment` representation for this file type.

```{r}
fra <- import(tfr)
fra
```

Similar operations to those used with SummarizedExperiment are supported. For
example, the genomic ranges can be displayed via `rowRanges`:

```{r}
rowRanges(fra)
```

# Session Information

```{r}
sessionInfo()
```

