---
title: "TENxIO: Import Single Cell Data Files"
author: "Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TENxIO Quick Start Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
Package: TENxIO
---

<!-- badges: start -->
<!-- badges: end -->

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction

`TENxIO` allows users to import 10X pipeline files into known Bioconductor
classes. The package is not comprehensive, there are files that are not
supported. It currently does not support Visium datasets. It does replace
some functionality in `DropletUtils`. If you would like a file format to be
supported. Please open an issue at https://github.com/waldronlab/TENxIO.

# Installation

```{r,eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("waldronlab/TENxIO")
```

# Load the package

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(TENxIO)
```

# Description

`TENxIO` offers an set of classes that allow users to easily work with files
typically obtained from the 10X Genomics website. These are commonly outputs
of the Cell Ranger pipeline.

# Supported Formats

| **Extension**       | **Class**     | **Imported as**      |
|---------------------|---------------|----------------------|
| .h5                 | TENxH5        | SingleCellExperiment |
| .mtx / .mtx.gz      | TENxMTX       | dgCMatrix            |
| .tar.gz             | TENxFileList  | SingleCellExperiment |
| peak_annotation.tsv | TENxPeaks     | GRanges              |
| fragments.tsv.gz    | TENxFragments | RaggedExperiment     |
| .tsv / .tsv.gz      | TSVFile\*     | tibble               |

**Note** (\*). The `TSVFile` class is used internally and not exported.

# TENxFile

The `TENxFile` class is the catch-all class superclass that allows transition
to subclasses pertinent to specific files. It inherits from the `BiocFile`
class and allows for easy dispatching `import` methods.

```{r}
showClass("TENxFile")
```

## `ExperimentHub` resources

`TENxFile` can handle resources from `ExperimentHub` with careful inputs. 
For example, one can import a `TENxBrainData` dataset via the appropriate
`ExperimentHub` identifier (`EH1039`):

```{r}
hub <- ExperimentHub::ExperimentHub()
hub["EH1039"]
```

Currently, `ExperimentHub` resources do not have an extension and it is best to
provide that to the `TENxFile` constructor function.

```{r,eval=FALSE}
fname <- hub[["EH1039"]]
TENxFile(fname, extension = "h5", group = "mm10", version = "2")
```

Note. `EH1039` is a large ~ 4GB file.

## TENxH5

`TENxIO` mainly supports version 3 and 2 type of H5 files. These are files
with specific groups and names as seen in `h5.version.map`, an internal
`data.frame` map that guides the import operations.

```{r}
TENxIO:::h5.version.map
```

In the case that, there is a file without genomic coordinate information, the
constructor function can take an `NA_character_` input for the `ranges`
argument.

The `TENxH5` constructor function can be used on either version of these H5
files. In this example, we use a subset of the PBMC granulocyte
H5 file obtained from the [10X website](https://cf.10xgenomics.com/samples/cell-arc/2.0.0/pbmc_granulocyte_sorted_3k/pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5).

```{r}
h5f <- system.file(
    "extdata", "pbmc_granulocyte_ff_bc_ex.h5",
    package = "TENxIO", mustWork = TRUE
)
library(rhdf5)
h5ls(h5f)
```

Note. The `h5ls` function gives you an overview of the structure of the file.
It matches version 3 in our version map. 

The show method gives an overview of the data components in the file:

```{r}
con <- TENxH5(h5f)
con
```

### import TENxH5 method

We can simply use the import method to convert the file representation to
a Bioconductor class representation, typically a `SingleCellExperiment`.

```{r}
import(con)
```

**Note**. Although the main representation in the package is
`SingleCellExperiment`, there could be a need for alternative data class
representations of the data. The `projection` field in the `TENxH5` show method
is an initial attempt to allow alternative representations.

# TENxMTX

Matrix Market formats are also supported (`.mtx` extension). These are typically
imported as SummarizedExperiment as they usually contain count data.

```{r}
mtxf <- system.file(
    "extdata", "pbmc_3k_ff_bc_ex.mtx",
    package = "TENxIO", mustWork = TRUE
)
con <- TENxMTX(mtxf)
con
import(con)
```

# TENxFileList

The `TENxFileList` class easily allows importing multiple files within a
`tar.gz` archive. The `untar` function can list the files compressed within the
tarball.

```{r}
fl <- system.file(
    "extdata", "pbmc_granulocyte_sorted_3k_ff_bc_ex_matrix.tar.gz",
    package = "TENxIO", mustWork = TRUE
)
untar(fl, list = TRUE)
```

Using a similar import process across all file types, one can easily obtain
a Bioconductor representation that is ready for analysis. `TENxFileList` can be
imported to `SingleCellExperiment`.

```{r}
con <- TENxFileList(fl)
import(con)
```

# TENxPeaks

Peak files can be handled with the `TENxPeaks` class. These files are usually
named `*peak_annotation` files with a `.tsv` extension. Peak files are
represented as `GRanges`.

```{r}
pfl <- system.file(
    "extdata", "pbmc_granulocyte_sorted_3k_ex_atac_peak_annotation.tsv",
    package = "TENxIO", mustWork = TRUE
)
tenxp <- TENxPeaks(pfl)
peak_anno <- import(tenxp)
peak_anno
```

# TENxFragments

Fragment files are quite large and we make use of the `Rsamtools` package to
import them with the `yieldSize` parameter. By default, we use a `yieldSize` of
200.

```{r}
fr <- system.file(
    "extdata", "pbmc_3k_atac_ex_fragments.tsv.gz",
    package = "TENxIO", mustWork = TRUE
)
```

Internally, we use the `TabixFile` constructor function to work with indexed
`tsv.gz` files.

**Note**. A warning is emitted whenever a `yieldSize` parameter is not set.

```{r}
tfr <- TENxFragments(fr)
tfr
```

Because there may be a variable number of fragments per barcode, we use a
`RaggedExperiment` representation for this file type.

```{r}
fra <- import(tfr)
fra
```

Similar operations to those used with SummarizedExperiment are supported. For
example, the genomic ranges can be displayed via `rowRanges`:

```{r}
rowRanges(fra)
```

# Session Information

```{r}
sessionInfo()
```

